---
title: "BikeShare_Success"
output:
  pdf_document: default
  html_document: default
date: "2022-11-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'D:/DataScience/DS_Project/DA_BikeShareSuccess/Data')
```

# Increase subscription rate for bike-share company

1. Objective: Design marketing strategies aimed at converting casual riders into annual members. 

2. Tasks: 
* Understand how annual members and casual riders differ
* Why casual riders would buy a membership
* Propose marketing strategy to maximize the subscription rate
* Explain how digital media could affect subscription rate

3. Provided data: monthly trip data from Divvy, a bike-share company in Chicago, from October 2021 to October 2022. 
* Source: https://ride.divvybikes.com/system-data
* This is part of the capstone project for Google Analytics Certificate

```{r}
## Import required libraries
library(tidyverse)
library(ggplot2)
library(lubridate)
library(skimr)
```

Check the current directory:
```{r}
getwd()
```
#### Import daily trip data for every month since October 2021-October 2022
```{r}
ldf <- list()  # creates a list
listcsv <- list.files(pattern = "*data.csv") #List all the data file in the folder
listcsv
```

```{r}
## Read monthly data set:
for (k in 1:length(listcsv)){
  ldf[[k]] <- read.csv(listcsv[k])
  
}
#View a sample of 1 dataset:
glimpse(ldf[[1]]) 

```

#### Wrangle data and combine into a single file:

Check the consistency of all column names in the data:
```{r}
colnames_df = colnames(ldf[[1]])
for (f in 1:length(ldf)){
  colnames_f= colnames(ldf[[f]])
  if (identical(colnames_df,colnames_f) != TRUE){
    cat('Column names in file',as.character(f),'is different\n')
    print(colnames_f)
  } else{
    cat('Column names in file',as.character(f),'is consistent. No change was made\n')
  }
}
```
#### Inspect several dataframe and look for incongruencies:

```{r}
str(ldf[[1]])
str(ldf[[7]])
str(ldf[[12]])
```

#### Combine all data together with the same features:

```{r}
all_trips <- bind_rows(ldf)
```

#### Clean data:

View data:
```{r}
colnames(all_trips)#List of all column name
```
```{r}
head(all_trips) #see the first 6 rows of the data frame
```
```{r}
str(all_trips) #See list of columns and data type
```
```{r}
summary(all_trips)#Get a statistical summary in data
```
##### Filter all_trips data:
Add column for ride_length and convert to numerical type
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
all_trips$ride_length <-as.numeric(all_trips$ride_length)
is.numeric(all_trips$ride_length)
```
Add column for month, year, week day, and day of the month
Filter any trip that has the ride length > 0
```{r}
all_trips_v1 <- all_trips %>% 
                    mutate (
                      ride_date = date(started_at),
                      month = month(ride_date),
                      year = year(ride_date),
                      week_day = wday(ride_date),
                      day_month = day(ride_date)) %>% 
                    filter(ride_length >0) 
```

Check if there is any ride_length below 0 or station name as HQ QR
```{r}
#Check if there is any ride_length below 0 or station name as HQ QR
all_trips_v1 %>% 
  filter(ride_length <0)
```


#### Conduct Analysis

##### Descriptive analysis on ride_length in hours:

```{r}
#Descriptive analysis on ride_length in hours:
all_trips_v1 %>% 
  group_by(member_casual) %>% 
  summarize(average_ride = mean(ride_length)/3600,
            midpoint = median(ride_length)/3600,
            longest_ride = max(ride_length)/3600,
            shortest_ride = min(ride_length)/3600)
```

The spread of the ride duration is very large and more than what the normal trips are, which could be due to several reasons: 
1. Error in the recorded when bike was taken away for quality check
2. Error in docking the bike when returning the bike to the station

Research on Divvy website: the membership perks will include unlimited free ride for 45 minutes. 
Thus, since we will only focus on the analyzing data with short trips of less than or equal to 60 minutes and greater than or equal to 10 minutes.
Only electric bikes and classic bikes will be included in the analysis.

##### Filter the data for short trips less than 45 minutes and greater than 10 seconds
```{r}
short_ride <- all_trips_v1 %>% 
            filter(ride_length<=60*60,
                   ride_length>=10*60,
                   rideable_type !='docked_bike')  #filter data with rent length below 45 minute
str(short_ride)
```

Remap values in member_casual and week day for clearer context
```{r}
short_ride<-short_ride %>% 
  mutate(
    member_casual= recode(member_casual,'member'='Subscriber','casual'='Non-subscriber'),
    week_day= recode(week_day,'1'='Sunday','2'='Monday','3'='Tuesday','4'='Wednesday','5'='Thursday','6'='Friday','7'='Saturday')                     
  )
str(short_ride)
```

##### Descriptive analysis on ride_length in hours:
```{r}
#Descriptive analysis on ride_length in hours:
short_ride %>% 
  group_by(member_casual) %>% 
  summarize(average_ride = mean(ride_length)/3600,
            midpoint = median(ride_length)/3600,
            longest_ride = max(ride_length)/3600,
            shortest_ride = min(ride_length)/3600)
```

```{r}
#plot histogram
ggplot(data = short_ride) +  # call the data
  geom_histogram(mapping= aes(x= ride_length,color=rideable_type,fill=rideable_type),alpha=0.5) +
  facet_grid(~member_casual)+
  labs(title="Ride length histogram plot",x="Length(s)", y = "Count")+
  theme_classic()+
  theme(legend.position="top")+
  guides(fill=guide_legend(title="Type"),color=guide_legend(title="Type")) 
```

Non subscriber and Subscriber groups have similar distribution. But the amount of non-subscribers are less than subscribers group. 
Classic bike is more popular among the subscribers and non-subscibers, which could be due to the preferences from the customers or the amount of electric bikes available versus the classic bikes. 

```{r}
ggplot(short_ride, aes(x=ride_length,color=member_casual)) +
  stat_ecdf(geom = "step")+
  labs(title="Emperical cumulative distribution ride length plot",x="Length(s)", y = "CDF")+
  facet_grid(~rideable_type)+
  theme_classic()+
  theme(legend.position="top")
```

Generally, the rent duration of non-subscribers group is longer than the subscribers group, which is quite surprising.

##### Find pattern according to year, month and day in the week:
```{r}
# See average ride time by day for member vs casual user:
average_week <- short_ride %>% 
                group_by(member_casual,week_day) %>% 
                summarize(average_length_per_trip = mean(ride_length))
average_week
```

```{r}
ggplot(data=average_week, aes(x=week_day, y=average_length_per_trip,group=member_casual)) +
  geom_line(aes(linetype=member_casual))+
  geom_point(aes(shape=member_casual))+
  labs(title="Average rent length for days in a week",x="Day", y = "Average length (s)")+
  theme_classic()+
  theme(legend.position="top")

```
```{r}
# See average ride time by month for member vs casual user:
average_month <- short_ride %>% 
                group_by(member_casual,month) %>% 
                summarize(average_length_per_trip = mean(ride_length),total = n()) 
                
average_month
```

```{r}
ggplot(data=average_month, aes(x=month, y=average_length_per_trip,group=member_casual,size())) +
  geom_line(aes(linetype=member_casual))+
  geom_point(aes(shape=member_casual))+
  labs(title="Average monthly rent length",x="Month", y = "Average length (s)")+
  theme_classic()+
  theme(legend.position="top")

```

From the calculated average daily and monthly rental length, it is concluded that non-subscribers will have a longer rental duration than the subscribers group. The peak season will start in April and end in October when the weather gets colder especially in Chicago. The rental length is also increased in the weekends (saturday and Sunday) since people will use more bikes to commute and for recreational purposes.

```{r}
ggplot(data=average_month)+
  geom_line(mapping=aes(x=month, y = total,linetype=member_casual))+
  labs(title="",x="Month", y = "Number of trips")+
  theme_classic()+
  theme(legend.position="top")


```

The total number of trips in subscribers group increases during the spring and summer and decreases during the fall and winter time.
Similar trend is found for the non-subscribers group however, the total number of trips is less than the subscribers groups.

### Proposals to influence casual riders to become members:
##### Proposal 1: Increase the number of trips taken by casual riders 
Offer some partial discount for bike rental in the off-peak days such as Monday, Tuesday to maximize the profit obtained during the off-peak days while promoting the bike-sharing culture among casual riders

##### Proposal 2: Improve bike-sharing app with added feature to broadcast the benefit of using bike-sharing services and cost benefit for membership program
*Adding a some features in the app that allow user to track the total number of trips, cost per trips, and carbon footprint based on the number of trips and duration, the cost saving strategy with membership account.
*Adding an alert feature when the trip to casual users who frequently rent 30 minutes about the membership options and its benefit.


##### Proposal 3: Marketing campaign in Spring
Run a marketing campaign to advertise the membership program starting from March to May before the peak season for bike-sharing services


### Save data

```{r}
write_csv(short_ride, "short_ride_202110_202210.csv")
write_tsv(average_week, "average_week_ride_202110_202210.csv")
write_tsv(average_month, "average_month_ride_202110_202210.csv")
```

